### 2022-06-15 paper.tex
lots of minor changes, last pass

\usepackage[a4paper,hmargin=1.5cm,vmargin=2cm]{geometry}
\setlength{\columnsep}{0.7cm}
%\usepackage{microtype} % nice typesetting  % doesn't work for me :-( -S
\usepackage{amsmath}
\usepackage{graphicx}

@@

\begin{minipage}{0.7\textwidth}
\small
\textit{Abstract} -- Single-cell-bisulfite sequencing (scBS) allows for assessing DNA methylation with single-basepair and single-cell resolution. Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability. Typically, this is simply done by dividing the genome into large tiles and averaging in these. We show that this coarse-graining dilutes signal, and suggest improved approaches to determine more informative regions quantify methylation in, and a better quantitation method than simple averaging. We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number. We also show that our method of finding variably methylated regions finds biologically meaningful loci. Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality required to work with scBS data.
\end{minipage}
\vspace{6.5ex}

@@

Here, we discuss bisulfite sequencing, a method to study DNA methylation. In brief, DNA is treated with bisulfite, which converts unmethylated cytosines to uracils which are read as thymine in subsequent PCR, while methylated cytosines are protected from conversion. After sequencing, these conversions allow to determine the methylation status of all cytosines covered by reads \citep{Frommer_1992}. Bisulfite sequencing can also be performed with single-cell resolution \citep{Smallwood_2014} and even in parallel with scRNA-seq \citep{Clark2018}.

In the present paper, we discuss strategies to analyze scBS data, suggest improvement to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.

@@

\begin{figure*}[t]
    \hspace{2.5cm}A\\[-7ex]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_A.png}\\
    \end{center}
    \hspace{2.5cm}B\\[-3ex]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_B.png}
    \end{center}
    \caption{\small \textbf{Effect of read position:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified. Two cells cover differing parts of the interval with one read each. If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells. \textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained. To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
    \label{fig:smoothres}
\end{figure*}

The standard approach to analyse scBS-Seq data is based on methodology developed for the analysis of single-cell RNA-Seq data. Therefore, we start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analysed, before we discuss scBS data analysis.

The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. A first goal is usually to assigning cell types or states to cells. To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on very many genes. Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score. Hence, the PCA space representation is suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

This procedure is commonly adapted when working with single-cell DNA-methylation data, so that established methods used in scRNA-Seq can be used.
The task is therefore how to construct a suitable matrix as input for the PCA. A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each tile. To this end, we identify all CpG sites in the tile for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell. Then we denote as average DNA methylation of this tile in a given cell the proportion of observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A). This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (methylation fractions) between 0 and 1. This matrix is now subjected to PCA. After the PCA, one can proceed using approaches known from scRNA-seq.

@@

We first discuss the task of quantifying the level of methylation in a given, fixed, genomic interval. In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for each of the two cells shown. The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider that cell as having stronger methylation in the interval. However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval. Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.

Therefore, we propose to first obtain, for each CpG position, an average of the methylation across all cells and, then quantify for each individual cell its deviation from this average. In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, and the red vertical lines show an individual cell's deviation from the ensemble average. We take the lengths of the red lines as signed values (``residual''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG), and take the average over the residual for all the CpGs in the interval that are covered by reads from this cell. In this average, we perform shrinkage towards zero via a pseudocount (to trade bias for variance, see Methods for detail) in order to dampen the signal in cells with low coverage of the interval.

The average thus obtained is what we use to quantify this cell's (relative) methylation in the interval. For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for a PCA. The signal to noise ratio in this matrix will be better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cells' covered CpG sites in a region. The reason for this is that we reduce the variation in situation as the one depicted in Figure \ref{fig:smoothres}A, where the methylation of the reads might differ strongly even though there is no actual evidence for a difference between the two cells.

@@

To demonstrate the value of our proposed improvements, we obtained two previously published scBS data sets and re-analysed them.

@@

The first data set is a single-cell multi-omics data set generated by \citet{argelaguet2019gastru}. In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5--E7.5) as well as single-cell transcriptomes from the same cells. In one of their analyses, the authors calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development. The authors identified several genes where promoter methylation predicts gene expression, but obtained low correlations for the vast majority of genes. We reasoned that VMRs obtained with the previously described sliding window approach might be a better predictor of gene expression.

In Figure \ref{figure:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with \texttt{scbs scan}.
This analysis revealed a number of genes whose expression does not correlate with promoter methylation, but correlates with VMR methylation.
This effect is also observed when using our proposed measure of methylation, the shrunken mean of residuals (Figure \ref{figure:correlation}B).

\begin{figure}
    \begin{center}
    A\hspace{.9\columnwidth}~\\
    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
    B\hspace{.9\columnwidth}~\\
    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
    C\hspace{.9\columnwidth}~\\
    \includegraphics[width=.95\columnwidth]{part_leonie_git/leonie_plots/comparison_meth_RNA.pdf}
    \end{center}
    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
    \textbf{(A)} Correlation of each gene's RNA expression values to the the methylation fractions that were measured at promoters (y-axis), or to methylation fractions of the nearest VMR (x-axis). Promoters are defined as intervals \textpm2~kb around the TSS.
    \textbf{(B)} Same as in (A), but using mean shrunken residuals as a measure of methylation.
    \textbf{(C)} Mean smoothed methylation near the gene Ube2l6 (ENSMUSG00000027078, marked in yellow in A \& B). Cells are assigned to 4 groups based on Ube2l6 expression (group 0: cells with no Ube2l6 expression, groups 1-3: cells that express Ube2l6, divided into three equally large groups with group 3 having the highest expression). The mean methylation level of each group was smoothed with a tricube kernel of bandwidth 2000.
    Data from \citet{argelaguet2019gastru}.}
    \label{figure:correlation}
\end{figure}

To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A,B).
Grouping all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.

@@

\begin{figure*}
	\begin{center}
		A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
		\hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
		\hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
		C\hspace*{.7\textwidth}~\\
		\hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
		D\hspace*{.7\textwidth}~\\
		\hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
		\caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}. \textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic regions (VMRs or 100~kb tiles) as input to PCA. Exemplary UMAPs for encircled analyses are shown in (C). \textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells. Exemplary UMAPs for analyses are shown in (D). \textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb tiles (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fraction of 100~kb tiles (left) or mean shrunken residuals of VMRs (right). Data from \citet{luo2017single}.}
		\label{figure:score}
	\end{center}
\end{figure*}

To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}. This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or mean shrunken residuals as a measure of methylation. 
For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 high-coverage regions that were selected for PCA.
Specifically, we ranked regions by their number of coverging cells (i.e., cells with at least on read in the region) and selected those with the highest number.

@@

Since scBS protocols is still rather costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
Here, our approach keeps its good score, while the simple approach falls behind (Figure \ref{figure:score}B).

@@

\texttt{scbs prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.~e. the number of CpG sites that have sequencing coverage. These summary statistics can be used to detect cells with poor quality. The quality of each single cell methylome can furthermore be inspected with \texttt{scbs profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS) at single-cell resolution. The TSS profile is a useful quality control plot since methylation shows a characteristic dip roughly \textpm1~kb around the transcription start site in mammalian genomes. Cells that do not show this pattern, or cells with few observed CpG sites, may then be discarded with \texttt{scbs filter}.

After quality control, the user has access to the genome-wide VMR detection approach described above via \texttt{scbs scan}. This produces a BED file that lists the genome coordinates of VMRs. To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells. The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells. We note that both \texttt{matrix} and \texttt{profile} accept any valid BED-file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells. The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

@@

Here, we have proposed an improved strategy to preprocess single-cell bisulfite sequencing data.
Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a reads local context rather than just treating all reads in an interval alike
Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
By reanalyzing two published datasets we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data. As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for smaller data sets with limited numbers of cells. In addition to this improvement of global signal-to-noise ratio (with regards to cell dissimilarity quantification), we also show that the determination of VMRs improve biological interpretability by pinpointing the regions that actually change and hence have a putative regulatory role.

Furthermore, we have presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms. The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting, thus allowing the user to use, e.g., well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses etc. By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.

@@

In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and also completes a gap in existing tool chains.

@@

We write $x_{ij}$ for the methylation status of CpG $i$ in cell $j$. The index $i$ runs over all CpG positions present in the genome, the index $j$ over all cells in the assay. We write $x_{ij}=0$ if position $i$ was found to be unmethylated in cell $j$ by bisulfite sequencing , $x_{ij}=1$ if it was methylated, and $x_{ij}=\text{NA}$ if position $i$ was not covered by reads from cell $j$ and the methylation status is therefore not available ("NA"),

@@

If multiple reads with the same cell barcode cover the positions, these will typically be PCR duplicates of each other and hence agree. Of course, the two alleles of a CpG may differ in their methylation status and while it is in principle possible that one obtains reads stemming the same position on both the paternal and the maternal chromosomes of the same cell, but this is so unlikely that we can ignore the case. Hence, whenever Bismarck reports multiple reads covering the same position in the same cell, we set $x_{ij}$ to 0 or 1 following the majority of reads and put $x_{ij}=\text{NA}$ in case of a tie.

@@

\subsection{Finding variably methylated regions (VMRs)}

@@

Given a set, $\mathcal{V}=\{I^\text{v}_1,I^\text{v}_2,\dots\}$, of intervals corresponding to VMRs, we get a relative methylation fraction $r_{ij}$ for each VMR $I^\text{v}_i$ and each cell $j$ from Eq.\ ()\ref{avgres}). The matrix thus obtained can then be centered and used as input for a PCA. If we calculate the top $R$ principal components, we thus obtain for each cell $j$ an $R$-dimensional principal component vector $\mathbf{x}^\text{P}_j$. For any two cells $j$, $j'$, we use the Euclidean distance $\|\mathbf{x}^\text{P}_j - \mathbf{x}^\text{P}_{j'}\|$ as the measure of dissimilarity of the two cells.
Thus, the matrix of PC scores can be used as input to dimension reduction methods like t-SNE or UMAP, and to clustering methods like the Louvain or Leiden algorithm, which require such a matrix as input to the approximate mnearest neighbor (ANN) finding algorithm that is their first step.

>>>>

\usepackage[a4paper,hmargin=1.5cm,vmargin=2cm]{geometry}
\setlength{\columnsep}{0.7cm}
\usepackage{microtype} % nice typesetting  % doesn't work for me :-( -S
\usepackage{amsmath}
\usepackage{graphicx}

@@

\begin{minipage}{0.7\textwidth}
\small
\textit{Abstract} -- Single-cell bisulfite sequencing (scBS) allows for assessing DNA methylation at single-base pair and single-cell resolution. Analysing the big data sets thus obtained requires preprocessing to reduce data size, improve signal over noise, and provide interpretability. Typically, this is simply done by dividing the genome into large tiles and averaging in these. We show that this coarse-graining dilutes signal, and suggest improved approaches to determine more informative regions to quantify methylation in, and a better quantitation method than simple averaging. We demonstrate how our approach enables better distinguishing of cell types and reduces demands on cell number. We also show that our method of finding variably methylated regions finds biologically meaningful loci. Finally, we present a software tool, \texttt{scbs}, that implements this approach and offers further functionality required to work with scBS data.
\end{minipage}
\vspace{6.5ex}

@@

Here, we discuss bisulfite sequencing, a method to study DNA methylation. In brief, DNA is treated with bisulfite, which converts unmethylated cytosines to uracils which are read as thymine in subsequent PCR, while methylated cytosines are protected from conversion. After sequencing, these conversions allow to determine the methylation status of all cytosines covered by reads \citep{Frommer_1992}. Bisulfite sequencing can also be performed at single-cell resolution \citep{Smallwood_2014} and even in parallel with scRNA-seq \citep{scMTseq,Clark2018}.

In the present paper, we discuss strategies to analyze scBS data, suggest improvements to current approaches, show their value in a benchmark, and present software to perform the improved analysis methods.

@@

\begin{figure*}[t]
    \hspace{2.5cm}A\\[-7ex]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_A.png}\\
    \end{center}
    \hspace{2.5cm}B\\[-3ex]
    \begin{center}
        \includegraphics[width=0.7\textwidth]{figures/Fig_residuals_B.png}
    \end{center}
    \caption{\small \textbf{Improved quantification of DNA methylation in genomic intervals:} \textbf{(A)} Depicted is a genomic interval along a chromosome, for which DNA methylation is to be quantified. Two cells cover differing parts of the interval with one read each. If one simply counts for each cell which fraction of the covered CpG sites are methylated, one obtains very different values for the two cells. \textbf{(B)} By averaging each CpG site's methylation over all cells and subsequent smoothing, the thick black ``average methylation curve'' is obtained. To quantify the methylation of cell 1 from (A) relative to this average over all cells, we propose to use the cell's residuals to the smoothed curve (lengths of the vertical red lines) and take their average, counting residuals of methylated CpGs positive and residuals of unmethylated CpGs negative.}
    \label{fig:smoothres}
\end{figure*}

The standard approach to analyze scBS-Seq data is based on methodology developed for the analysis of single-cell RNA-Seq data. Therefore, we start by briefly reviewing how single-cell \emph{RNA}-Seq data is commonly analyzed, before we discuss scBS data analysis.

The starting point in most scRNA-Seq analyses is a matrix of UMI counts (i.e. of counts of distinct RNA molecules), one row for each cell and one column for each gene. A first goal is usually to assign cell types or states to cells. To this end, one needs to establish which cells are similar to each other, i.e., quantify the distance (i.e., dissimilarity) between any two given cells' transcriptional profile. A standard approach, used with minor variation in virtually all recent research and automated by popular software such as Seurat \citep{Hao_2021} or ScanPy \citep{Wolf_2018}, is as follows: One first accounts for cell-to-cell variation in sequencing depth by dividing each UMI count by the respective cell's total UMI count, then transforms to a homoskedastic scale by taking the logarithm. In order to avoid matrix elements with zero count to be transformed to minus infinity, one typically adds a very small ``pseudocount'' (often $10^{-4}$) to the normalized fractions before taking the logarithm. Now, one could use Euclidean distances of these vectors of logarithmized fractions as dissimilarity score. However, these scores would be exceedingly noisy due to the strong Poisson noise introduced by the many genes with very low counts. Therefore, one performs a principal component analysis (PCA), keeping only the top few (typically, 20 to 50) components. As Poisson noise is uncorrelated between genes, it will average out in the top principal components, as these are all linear combinations with weight on very many genes. Therefore, Euclidean distances between these ``PCA space'' vectors provide a robust dissimilarity score. Hence, the PCA space representation is suitable as input to methods like t-SNE and UMAP, which provide a two-dimensional representation of the data, or to methods for clustering (assigning cells to groups by similarity) and trajectory finding (identifying elongated manifolds in PCA space and assigning cells to quasi-1D positions along them).

This procedure is commonly adapted when working with single-cell DNA methylation data, so that established methods used in scRNA-Seq can be used.
The task is therefore how to construct a suitable matrix as input for PCA. A simple and common approach, used for instance by \citet{luo2017single}, is to divide the genome into tiles of, say, 100~kb size, and calculate for each cell the average methylation of the DNA within each tile. To this end, we identify all CpG sites in the tile for which we have coverage with at least one read and can hence call the CpG to be either unmethylated or methylated in the cell. Then we denote as average DNA methylation of this tile in a given cell the proportion of observed CpG sites in the tile that were found to be methylated (Figure \ref{fig:smoothres}A). This yields a matrix, with one row for each cell and one column for each genomic tile, comprising numbers (methylation fractions) between 0 and 1. This matrix is now subjected to PCA. After PCA, one can proceed using approaches known from scRNA-seq.

@@

We first discuss the task of quantifying the level of methylation in a given, fixed, genomic interval. In Figure \ref{fig:smoothres}A, such an interval is covered by a single read for two of the three cells shown. The read from cell 2 shows much more methylation than the read from cell 1, and the standard analysis would therefore consider that cell as having stronger methylation in the interval. However, given that the two reads agree wherever they overlap, a more parsimonious interpretation would be that the cells do not show difference in methylation within the interval. Rather, both cells, and similarly maybe most other cells, might have stronger methylation in the left third of the interval than in the middle one.

Therefore, we propose to first obtain, for each CpG position, an average of the methylation across all cells, and then quantify for each individual cell its deviation from this average. In Figure \ref{fig:smoothres}B, the curved line depicts such an average over all cells, and the red vertical lines show an individual cell's deviation from the ensemble average. We take the lengths of the red lines as signed values (``residual''), positive for lines extending upwards from the curve (methylated CpG) and negative for lines extending downwards (unmethylated CpG), and take the average over the residual for all the CpGs in the interval that are covered by reads from this cell. In this average, we perform shrinkage towards zero via a pseudocount (to trade bias for variance, see Methods for detail) in order to dampen the signal in cells with low coverage of the interval.

The average thus obtained is what we use to quantify this cell's (relative) methylation in the interval. For a genome tiled into such intervals, we thus obtain a matrix, one row per cell, one column per interval, that can be used for downstream analysis, e.g., as input for PCA. The signal-to-noise ratio in this matrix will be better than in a matrix obtained by simply averaging absolute methylation (0 or 1) over all the cells' covered CpG sites in a region. The reason for this is that we reduce the variation in situations as the one depicted in Figure \ref{fig:smoothres}A, where the methylation of the reads might differ strongly even though there is no actual evidence for a difference between the two cells.

@@

To demonstrate the value of our proposed improvements, we obtained two previously published scBS data sets and re-analyzed them.

@@

The first data set used for benchmarking is a single-cell multi-omics data set generated by \citet{argelaguet2019gastru}. In their work, the authors obtained single-cell methylomes of mouse embryo cells across different stages of development (E4.5--E7.5) as well as single-cell transcriptomes from the same cells. In one of their analyses, the authors calculated the correlation between gene expression and promoter methylation to elucidate the implications of the epigenome on gene expression in development. The authors identified several genes where promoter methylation predicts gene expression, but obtained low correlations for the vast majority of genes. We reasoned that VMRs obtained with the previously described sliding window approach might be a better predictor of gene expression.

In Figure \ref{figure:correlation}A, we compare correlations of gene expression and promoter methylation to correlations of gene expression and methylation of the closest VMR identified with our approach.
This analysis revealed a number of genes whose expression does not correlate with promoter methylation, but correlates with VMR methylation.
This effect is also observed when using our proposed measure of methylation, the shrunken mean of residuals (Figure \ref{figure:correlation}B).

\begin{figure}
    \begin{center}
    A\hspace{.9\columnwidth}~\\
    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_methylation_fraction_4kbwindow.pdf} \\
    B\hspace{.9\columnwidth}~\\
    \includegraphics[width=.7\columnwidth]{part_leonie_git/leonie_plots/corr_shrunken_residual_4kbwindow.pdf} \\
    C\hspace{.9\columnwidth}~\\
    \includegraphics[width=.95\columnwidth]{part_leonie_git/leonie_plots/comparison_meth_RNA.pdf}
    \end{center}
    \caption{\small \textbf{Correlation of DNA methylation and gene expression}.
    \textbf{(A)} Correlation of each gene's RNA expression values to the the methylation fractions that were measured at promoters (y-axis), or to methylation fractions of the nearest VMR (x-axis). Promoters are defined as intervals \textpm2~kb around the TSS.
    \textbf{(B)} Same as in (A), but using mean shrunken residuals as a measure of methylation.
    \textbf{(C)} Mean smoothed methylation near the gene Ube2l6 (ENSMUSG00000027078, marked in orange in A \& B). Cells are assigned to 4 groups based on Ube2l6 expression (group 0: cells with no Ube2l6 expression, groups 1-3: cells that express Ube2l6, divided into three equally large groups with group 3 having the highest expression). The mean methylation level of each group was smoothed with a tricube kernel of bandwidth 2000.
    Data from \citet{argelaguet2019gastru}.}
    \label{figure:correlation}
\end{figure}

To illustrate this effect with an example, we picked a gene where VMR methylation is a much better predictor of gene expression than promoter methylation (Ube2l6, highlighted in orange in Figure \ref{figure:correlation}A,B).
Dividing all cells into four groups according to their level of Ube2l6 expression, it is clearly visible that the promoter of this gene is lowly methylated regardless of gene expression, while methylation at the nearest VMR is greatly decreased in cells with high Ube2l6 expression.

@@

\begin{figure*}
    \begin{center}
        A\hspace{.4\textwidth}B\hspace*{.3\textwidth}~\\
        \hspace{.3cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/complete_135k_12cm_log.pdf}
        \hspace{1cm}\includegraphics[width=.35\textwidth]{part_leonie_git/leonie_plots/cell500_20k_12cm_log.pdf}\\
        C\hspace*{.7\textwidth}~\\
        \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_fulldataset.pdf}\\
        D\hspace*{.7\textwidth}~\\
        \hspace{.3cm}\includegraphics[width=.75\textwidth]{part_leonie_git/leonie_plots/UMAP_reduceddataset.pdf}
        \caption{\small \textbf{Benchmark of our methods on single-cell methylomes of 16 neuronal subtypes}. \textbf{(A)} Nearest neighbor count score using four different method combinations and a variable number of genomic regions (VMRs or 100~kb tiles) as input to PCA. Exemplary UMAPs for encircled analyses are shown in (C). \textbf{(B)} Same benchmark as in (A), but for a smaller data set consisting of 500 randomly subsampled cells. Exemplary UMAPs for analyses are shown in (D). \textbf{(C)} Two exemplary UMAPs of the full data set and methylation fractions of 100~kb tiles (left) or mean shrunken residuals of VMRs (right) \textbf{(D)} Two exemplary UMAPs of the smaller data set and methylation fractions of 100~kb tiles (left) or mean shrunken residuals of VMRs (right). Data from \citet{luo2017single}.}
        \label{figure:score}
    \end{center}
\end{figure*}

To assess the impact of our methods on dimensionality reduction and the ability to distinguish different (sub)types of cells, we used a data set from \citet{luo2017single}. This data set consists of 3377 single-cell CpG methylomes from the murine frontal cortex, comprising 16 neuronal subtypes.
We subjected this data set to the standard scBS workflow: defining a set of genomic intervals, quantifying methylation in these intervals, and subjecting the resulting matrix to PCA.
We repeated this analysis several times, either using 100~kb tiles or VMRs, and either using methylation fractions or the shrunken mean of residuals as a measure of methylation. 
For both types of regions, we varied the number of input regions, ranging from 100 to 50,000 regions with the highest sequencing coverage that were selected for PCA.

@@

Since scBS protocols are still rather costly and labor-intensive, only few laboratories are currently able to obtain thousands of single cell methylomes.
To simulate a smaller data set, we subsampled the frontal cortex data set to 500 cells and repeated our analysis.
Here, our approach keeps its good score, while the simple approach falls behind (Figure \ref{figure:score}B).

@@

\texttt{scbs prepare} also computes a number of summary statistics for each cell, including the mean genome-wide methylation level and the number of observed CpG sites, i.e. the number of CpG sites that have sequencing coverage. These summary statistics can be used to detect cells with poor quality. The quality of each single cell methylome can furthermore be inspected with \texttt{scbs profile}, which computes the average methylation profile of a set of user-defined genomic regions such as transcription start sites (TSS) at single-cell resolution. The TSS profile is a useful quality control plot since methylation shows a characteristic dip roughly \textpm1~kb around the transcription start site in mammalian genomes. Cells that do not show this pattern, or cells with few observed CpG sites, may then be discarded with \texttt{scbs filter}.

After quality control, the user has access to the genome-wide VMR detection approach described earlier, using the \texttt{scbs scan} command. This produces a BED file that lists the genome coordinates of VMRs. To finally obtain a methylation matrix analogous to a scRNA-seq count matrix, this BED file can be used as input for \texttt{scbs matrix}, which quantifies methylation at genomic intervals in all cells. The command produces both the percentage of methylated CpG sites, as well as our proposed methylation measure (shrunken mean of the residuals) that is more robust to variation in sequencing coverage or stochastic differences in read position between cells. We note that both \texttt{matrix} and \texttt{profile} accept any valid BED file as input, which means that the user can quantify and visualize methylation at any genomic feature of interest, such as promoters, enhancers or transcription factor binding sites, in single cells. The obtained methylation matrix can then serve as input for common single-cell analysis methods such as dimensionality reduction and cell clustering.

@@

Here, we have proposed an improved strategy to preprocess single-cell bisulfite sequencing data.
Based on the observation that incomplete read coverage of a genomic interval can lead to inaccurate methylation estimates, we suggest a scoring scheme that is aware of a reads local context rather than just treating all reads in an interval alike.

Furthermore, we show a way to pinpoint minimal regions of high variability across cells, which we call variably methylated regions (VMRs).
By re-analyzing two published data sets, we could show that these improvements to data preprocessing help to increase signal and decrease noise, resulting in a more informative intermediate-dimensional representation of the data. As examples of practical benefits, we demonstrate that our preprocessing allows for better distinction of cell sub-types, especially for smaller data sets with a limited number of cells. In addition to this improvement of global signal-to-noise ratio (with regards to cell dissimilarity quantification), we also show that the determination of VMRs improve biological interpretability by pinpointing the regions that actually change and hence have a putative regulatory role.

We also presented and published an open source software tool, called \texttt{scbs}, that provides an easy-to-use implementation of the described algorithms. The package also offers functionality for all other steps of data preprocessing that are needed to get from the output of methylation callers such as Bismark and methylpy to a cell\texttimes region matrix that is suitable as input for standard analysis tools that work in a reduced-dimensionality setting, thus allowing the user to use, e.g., well established tools from scRNA-seq analysis, including visualization tools such as t-SNE and UMAP, Leiden/Louvain clustering, pseudotime trajectory analyses etc. By offering these functions, \texttt{scbs} bridges a gap in the chain of existing tools that so far hindered practitioners in their data analysis.

@@

In conclusion, we have presented powerful improvements to scBS data preprocessing and a software tool that implements these and bridges a gap in existing tool chains.

@@

We write $x_{ij}$ for the methylation status of CpG $i$ in cell $j$. The index $i$ runs over all CpG positions present in the genome, the index $j$ over all cells in the assay. We write $x_{ij}=0$ if position $i$ was found to be unmethylated in cell $j$ by bisulfite sequencing, $x_{ij}=1$ if it was methylated, and $x_{ij}=\text{NA}$ if position $i$ was not covered by reads from cell $j$ and the methylation status is therefore not available ("NA"),

@@

If multiple reads from the same cell cover a position, these will typically be PCR duplicates of each other and hence agree. Of course, the two alleles of a CpG may rarely differ in their methylation status. While it is in principle possible that one obtains discordant reads stemming from the same position on both the paternal and the maternal chromosomes of the same cell, this is so unlikely that we can ignore the case. Hence, whenever the methylation caller reports multiple reads covering the same position in the same cell, we set $x_{ij}$ to 0 or 1 whenever all reads agree.
When there is disagreement, we put $x_{ij}=\text{NA}$ by default, or optionally follow the majority of reads whenever possible.

@@

@@

Given a set, $\mathcal{V}=\{I^\text{v}_1,I^\text{v}_2,\dots\}$, of intervals corresponding to VMRs, we get a relative methylation fraction $r_{ij}$ for each VMR $I^\text{v}_i$ and each cell $j$ from Eq.\ (\ref{avgres}). The matrix thus obtained can then be centered and used as input for a PCA. If we calculate the top $R$ principal components, we thus obtain for each cell $j$ an $R$-dimensional principal component vector $\mathbf{x}^\text{P}_j$. For any two cells $j$, $j'$, we use the Euclidean distance $\|\mathbf{x}^\text{P}_j - \mathbf{x}^\text{P}_{j'}\|$ as the measure of dissimilarity of the two cells.
Thus, the matrix of PC scores can be used as input to dimension reduction methods like t-SNE or UMAP, and to clustering methods like the Louvain or Leiden algorithm, which require such a matrix as input to the approximate nearest neighbor (ANN) finding algorithm that is their first step.
